---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## akfingapdata

<!-- badges: start -->
<!-- badges: end -->

This pacakge provides functions to download [gap_product](https://afsc-gap-products.github.io/gap_products/) tables from the AKFIN database using web services. 

## Installation
``` {r, eval=FALSE}
# devtools::install_github("MattCallahan-NOAA/akfingapdata")
```

## Authentication

In order to use this package you must be authorized to view these data.
Please contact matt.callahan@noaa.gov for assistance setting up web service credentials.
Authorized users will be issued a secret text string in a .txt file, which will be in the format of 1234r..:abcd2.. 
The create_token function encrypts this secret string to authenticate access to these data.
The create_token function accepts a defined variable as the secret string, the filepath to the textfile, or a keyring package service input.
This token is valid for about ten minutes. If it expires you will get an error asking you to provide a valid token.
You must define the function output as token as in the example below. 

```{r message = FALSE}
library(akfingapdata)
library(tidyverse)
library(keyring)
library(kableExtra)

token<-create_token("akfin_secret") # keyring input
# token<-create_token("Callahan_token.txt") file version
```

## Metadata
A good first step is to download the column descriptions for these data.
The authors of these tables mercifully named columns in such a way that no column has different definitions in different tables, so one metadata table will suffice for all tables. 

```{r message = FALSE}
metadata<-get_gap_metadata_column()
head(metadata %>% dplyr::select(metadata_colname,metadata_colname_long, metadata_units)) %>%kable()

```

## Lookup tables

Some functions require survey_definition_id, species_code, area_id, start_year, and end_year as parameters.  
In order to correctly define these parameters, first download the species lookup table. 

```{r }
# download species_code
# The akfin_taxonomic_groups table supersedes the akfin_taxonomic_classification table, these tables have the same structure, except the groups table has a group column designating which group code a species corresponds to for species that have comps calculated at the group level rather than individually. 
taxa<- get_gap_taxonomic_groups()

# find species codes for shortraker rockfish
 taxa %>% filter(grepl("shortraker", tolower(common_name))) %>%
   dplyr::select(group_code, species_code, species_name, common_name) %>% kable()
```
The GAP_PRODUCTS data structure uses the survey_definition_id column to specify which survey (region) data belong to. 

```{r}
# The survey name can be found in the cruise table.
cruise <- get_gap_cruise()

cruise %>%
  group_by(survey_definition_id) %>% 
  summarize(survey_name = unique(survey_name)) %>% kable()
```

The design_year field notes the year of survey design. Biomass, agecomp, and sizecomp estimates are provided for one design year only. As of summer 2025, for all surveys except the GOA, indices use the most recent survey design year. The GOA survey design changed in 2025 and old estimates were not recalculated with the new design, therefore in the GOA 1990-2023 use the 2024 design year and 2025 onward use the 2025 design year. 

```{r}
# download survey design
survey_design<-get_gap_survey_design()

survey_design %>% 
  group_by(survey_definition_id) %>%
  reframe(design_year = unique(design_year)) %>% kable()
  
```


The get_gap_area() function provides the akfin_area table, which contains area_id descriptions. Biomass, sizecomp, and agecomp estimates are provided at the area_id level. Determining the correct area_id is one crux to replicate results from pre-gap_products tables. Area_ids are grouped by area types e.g "REGION", "REGULATORY AREA", "STRATUM". The akfin_area table has records for each design_year, so if joining the area descriptions to biomass estimates, design_year needs to be specified to avoid duplicates. 

```{r}
area <- get_gap_area()

unique(area$area_type)
```

The get_gap_stratum_groups() function shows which strata are in each AREA_ID (for non "STRATUM" area_ids).
```{r}
stratum_groups <- get_gap_stratum_groups()

# which stata are in the Western GOA?
area %>%
  filter(design_year ==2025,
         survey_definition_id==47,
         area_type=="REGULATORY AREA",
         area_name == "Western GOA"
         ) %>%
  dplyr::select(area_id)

stratum_groups %>%
  filter(design_year == 2025,,
         survey_definition_id==47,
         area_id == 805) %>% kable()
```


## Data tables
Each table has a separate function to pull data from it.
Agecomp, Biomass, Catch, CPUE, Length, Sizecomp, and Specimen tables have hundreds of thousands to over 100 million rows.
Parameters are built into the web services and functions to prevent enormous downloads.
Default parameters are as follows: survey_definition_id= 98 (Bering Sea), area_id = 1 (EBS Subarea 1), species_code=21740 (Pollock), start_year = 1990, and end_year = 3000.
Agecomp, biomass, and sizecomp are calculated at the area_id level and survey_definition_id and area_id are built into those tables.

```{r}
# Get GOA shortraker, agecomp, biomass, and sizecomp data 
goasr_biomass<-get_gap_biomass(survey_definition_id = 47,
                area_id = 99903,
                species_code = 30576,
                start_year=1990,
                end_year=2023)

goasr_sizecomp<-get_gap_sizecomp(survey_definition_id = 47,
                area_id = 99903,
                species_code = 30576,
                start_year=1990,
                end_year=2023)


head(get_gap_agecomp(survey_definition_id = 47,
                area_id = 99903,
                species_code = 30576,
                start_year=1990,
                end_year=2023))%>% kable()

```

Note that Eastern Bering Sea stratum level age compositions are calculated using "EBS STANDARD" and "EBS STANDARD PLUS NW" girds. The get_gap_agecomp() function has an optional area_id_footprint argument that specifies which footprint to use (default is "EBS STANDARD PLUS NW"). 

```{r}
unique(get_gap_agecomp(
  survey_definition_id = 98, area_id = 50,species_code = 21370,start_year=1990,end_year=2023,area_id_footprint = "EBS STANDARD PLUS NW")$area_id_footprint)

unique(get_gap_agecomp(
  survey_definition_id = 98, area_id = 50,species_code = 21370,start_year=1990,end_year=2023,area_id_footprint = "EBS STANDARD")$area_id_footprint)

```


Catch, CPUE, Length, and Specimen tables contain haul level data.
The web service joins the haul and cruises tables to allow querying by survey.

```{r}
# Get catch, cpue, length frequency, and specimens at a haul level for GOA shortraker

# The catch table only includes records where the species were caught
goasr_catch<-get_gap_catch(survey_definition_id = 47,
                species_code = 30576,
                start_year=1990,
                end_year=2023)

# The CPUE table includes zeros and thus has many more records
goasr_cpue<-get_gap_cpue(survey_definition_id = 47,
                species_code = 30576,
                start_year=1990,
                end_year=2023)

# The length table has ~4 million rows as of 2025 so loads may be slow. 
goasr_length<-get_gap_length(survey_definition_id = 47,
                species_code = 30576,
                start_year=1990,
                end_year=2023)


# The specimen table contains length, sex, weight, and age data for individual fish.
goasr_specimen<-get_gap_specimen(survey_definition_id = 47,
                species_code = 30576,
                start_year=1990,
                end_year=2023)

head(goasr_specimen)
```

To query more than one species (or area) at a time use the lapply function or a for loop. 

```{r}
# Get rougheye and shortraker biomass for the WGOA
taxa %>% filter(grepl("rougheye", common_name))

myspecies <- c(30050, 30051)

goa_resr_biomass<-lapply(myspecies, FUN = function(x) get_gap_biomass(
  survey_definition_id = 47,
  area_id = 805,
  species_code = x,
  start_year=2007,
  end_year=2023)) %>%
  bind_rows() %>%
  arrange(year)


head(goa_resr_biomass)

# trying the same thing with the other functions produces the same generic error as when the token is expired.

```

## Cruise and haul tables
Haul and cruise level details are available for download as well.
Download these tables in full and manipulate in R. 
```{r}
gap_haul<-get_gap_haul()

gap_cruise<-get_gap_cruise()

head(gap_haul)

```


## Split Fractions
This table shows what proportion of fish are in the eastern or western GOA. 
It is downloaded in its entirety.
```{r}
split_fractions<-get_gap_split_fractions()
```

